---
title: "PM566 Final Project"
author: "Christina Lin"
output:
  html_document:
    toc: yes
---

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
options(scipen = 1, digits = 2)
library(data.table)
library(tidyverse)
library(tidytext)
library(ggplot2)
library(tidyr)
library(tibble)
library(dplyr)
library(plotly)
library(DT)
library(knitr)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 9, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")
```

# Exploring Cytochrome P450 Enzymes Found in the Human Brain

## Introduction
This project is based on my PhD thesis exploring the enzymes in the brain that can produce the neurosteroid pregnenolone from cholesterol. Neurosteroids have important functions in the brain including neuroprotection, anti-inflammation, and modulation of behavior. In classical steroid-producing organs such as the adrenals, pregnenolone is metabolized from cholesterol by the cytochrome P450 (CYP450) enzyme CYP11A1 (see [Lin & Papadopoulos, 2021](https://pubmed.ncbi.nlm.nih.gov/34015388/) for review). However, CYP11A1 protein is difficult to detect in the brain and preliminary experiments have revealed that a potential alternate pathway not involving CYP11A1 is used by human brain cells to produce pregnenolone. Our data also suggests that this alternative pathway involves activity of CYP450s. Therefore, this project will analyze known CYP450s in the UniProt database as well as microarray data from the Allen Brain Atlas to answer 2 main questions: 1) Which CYP450 enzymes are expressed in the brain and what are their expression levels? 2) Which of those CYP450 enzymes are involved in cholesterol/steroid metabolism? 

<br>
<br>


## Methods
[List of CYP450s](https://www.uniprot.org/uniprot/?query=cytochrome%20p450&fil=organism%3A%22Homo%20sapiens%20(Human)%20%5B9606%5D%22%20AND%20reviewed%3Ayes&columns=id%2Centry%20name%2Cprotein%20names%2Cgenes%2Corganism%2Clength%2Cmass%2Ccomment(TISSUE%20SPECIFICITY)%2Ccomment(COFACTOR)%2Ccomment(FUNCTION)%2Ccomment(SUBCELLULAR%20LOCATION)%2Ccomment(PATHWAY)%2Csequence&sort=score) were obtained from the UniProt database by searching "cytochrome P450" and cleaned up to extract proteins of interest. To look for CYP450s that are expressed in the brains, terms such as "brain", "cerebellum", "cerebral",and "hippocampus" (i.e. common references to different parts of the brain) will be used to filter the observations.

<br>

The limitation of using UniProt is that it only shows information about proteins and it does not quantify expression levels. Some CYP450s that are expressed in the brain at low levels, such as CYP11A1, were also not recorded by UniProt. Therefore, we will use another database, Allen Brain Atlas, to look quantitatively at expression of CYP450s in the human brain. The term "CYP" was used to search for cytochrome P450s and the [results](https://human.brain-map.org/microarray/search/show?exact_match=false&search_term=CYP&search_type=gene&page_num=0) were downloaded for further analysis. CYP450 expression is represented in a log2 scale (i.e. a difference of 1 in expression means 2-fold change in RNA expression)

```{r, echo=FALSE}
cyp450 <- fread("data/UniProt_hCYP450s.csv")
setnames(cyp450, "Gene names", "Gene_name")
setnames(cyp450, "Protein names", "Protein_name")
setnames(cyp450, "Function [CC]", "Function")
setnames(cyp450, "Subcellular location [CC]", "Subcellular_location")
setnames(cyp450, "Tissue specificity", "Tissue_expression")
```

```{r, echo=FALSE}
cyp450$Gene_name <- stringr::str_extract(cyp450$Gene_name, "CYP[[:alnum:]]+")
start_rows <- nrow(cyp450)
cyp450 <- cyp450[!is.na(Gene_name),]
end_rows <- nrow(cyp450)
```

```{r, echo=FALSE}
cyp450$Mass <- stringr::str_remove_all(cyp450$Mass, ",")
cyp450$Mass <- as.integer(cyp450$Mass)
```

```{r, echo=FALSE}
cyp450 <- cyp450[Mass >= 35000,]
smallest <- cyp450[which.min(Mass),]
largest <- cyp450[which.max(Mass),]
```

```{r, echo=FALSE}
cyp450 <- cyp450[, Tissue_expression := str_to_lower(Tissue_expression)]
cyp450 <- cyp450[, Function := str_to_lower(Function)]
cyp450 <- cyp450[, Subcellular_location := str_to_lower(Subcellular_location)]
cyp450 <- cyp450[, Pathway := str_to_lower(Pathway)]
brain_cyp450 <- cyp450[grep("brain|cerebellum|cerebral|hippocampus", Tissue_expression),]
brain_cyp450 <- brain_cyp450[, Brain := as.factor(1)]
non_brain_cyp450 <- cyp450[!grep("brain|cerebellum|cerebral|hippocampus", Tissue_expression),]
non_brain_cyp450 <- non_brain_cyp450[, Brain := as.factor(0)]
```

```{r, echo=FALSE}
expressions <- fread("data/Expression.csv")
probes <- fread("data/Probes.csv")
tissues <- fread("data/Columns.csv")
probe_ids <- expressions[,1]
expressions <- expressions [, -1]
colnames(expressions) <- as.character(seq(1,ncol(expressions)))
row.names(expressions) = unlist(probe_ids)
expressions <- rownames_to_column(expressions, var = "id")
expressions <- expressions %>% pivot_longer(cols = 2:ncol(expressions), names_to = "tissue_id", values_to = "expression")
probes$id <- as.character(probes$id)
tissues <- cbind(tissue_id = as.character(seq(1, length(tissues$donor_id))),tissues)
microarray <- merge(probes,expressions, by = "id", all = TRUE)
microarray <- merge(microarray, tissues, by = "tissue_id", all.x = TRUE, all.y = FALSE)
microarray <- microarray[grep("CYP", `gene-symbol`)]
```



## Preliminary Results

The average mass of cyptochrome P450 enzymes found in humans is `r mean(cyp450$Mass)` Dalton and the average length is `r mean(cyp450$Length)` amino acids. The smallest CYP450 is `r smallest$Gene_name`, which is `r smallest$Mass` Dalton in mass and `r smallest$Length` amino acids in length. The largest CYP450 is `r largest$Gene_name`, which is `r largest$Mass` Dalton in mass and `r largest$Length` amino acids in length.


```{r, echo=FALSE}
ggplot(cyp450, mapping=aes(x = Mass, y = Length, color = Gene_name)) +
  geom_point() +
  xlab("Mass (Dalton)") +
  ylab("Length (amino acids)") +
  ggtitle("Mass and length of human CYP450s")
```

**Figure 1: Correlation between mass and length of human CYP450s.** As expected, there is a positive linear correlation between mass and length of CYP450 enzymes. 

<br>
<br>


Out of the `r nrow(cyp450)` CYP450 enzymes in humans, there are `r nrow(brain_cyp450)` CYP450s that have protein expression in the brain.

```{r, echo=FALSE}
brain_summary <- brain_cyp450[,.(Gene_name, Protein_name, Length, Mass)]
knitr::kable(brain_summary)
```
**Table 1: List of CYP450s expressed in the brain.** 

<br>
<br>


Next, the brain CYP450s proteins involved in cholesterol- or steroid-related functions or pathways will be examined. 

```{r, echo=FALSE}
chol_function <- brain_cyp450[grep("cholesterol|steroid|hormone", Function),]
chol_pathway <- brain_cyp450[grep("cholesterol|steroid|hormone", Pathway),]
```

CYP450s with cholesterol- and/or steroid-related functions are `r chol_function$Gene_name`. CYP450s involved in cholesterol- and/or steroid-related pathways are `r chol_pathway$Gene_name`. We are able to capture more enzymes when searching for CYP450s involved in cholesterol or steroid related functions. Of the `r nrow(brain_cyp450)` CYP450 enzymes expressed in the brain, `r nrow(chol_function)` of them have functions relating to cholesterol or steroid hormones. 
<br>
<br>


```{r, echo=FALSE}
chol_function[, ER := grepl("endoplasmic\\sreticulum",Subcellular_location)]
chol_function[, Mitochondria := grepl("mitochondria|mitochondrion", Subcellular_location)]
chol_function[, Cytoplasm := grepl("cytoplasm|cytoplasmic", Subcellular_location)]
chol_function[, Organelle := ifelse(ER&Mitochondria, "Both ER and Mitochondria", 
                                  ifelse(Mitochondria, "Mitochondria Only", "ER Only"))]

chol_function %>%
  plot_ly(x=~Organelle, y=~Mass, 
           type= 'scatter', mode = 'markers', color=~Gene_name, 
           hoverinfo='text',
           text=~Protein_name) %>%
 layout(title = "Subcellular Location of Cholesterol and Steroid Related Brain CYP450s",
                  yaxis = list(title = "Mass (Dalton)"), xaxis = list(title = "Subcellular Location"),
         hovermode = "compare")

```


**Figure 2: Subcellular location and masses of human CYP450s expressed in the brain that are involved in cholesterol and steroid functions.** The majority of these enzymes are located in the endoplasmic reticulum. CYP27A1 is the only one of these enzymes that is only located in the mitochondria while CYP1B1 is found in both the endoplasmic reticulum and mitochondria. The enzymes that localize to the mitochondria also appear to have higher masses than those that are only found in the ER. 

<br>
<br>


Therefore, the CYP450 enzymes of interest found using the UniProt database are listed in the table below. 
```{r, echo=FALSE}
interest <- chol_function[, .(Gene_name, Protein_name, Mass, Length, Organelle)]
knitr::kable(interest)
```
**Table 2: CYP450 Proteins Expressed in the Human Brain with Cholesterol and/or Steroid-related Functions.** 

<br>
<br>


However, the data from UniProt are not quantitative and the assays used to detect protein may not be sensitive enough to measure the expression of CYP450s in the brain, which can be bound to organelles or membranes. Therefore, the RNA expression for CYP450s were also examined and compared to CYP11A1, the enzyme whose activity we are trying to find an alternative pathway for. 

```{r, warning = FALSE, echo=FALSE}


ggplot(microarray, mapping=aes(x = top_level_structure_name, y= expression, color = `gene-symbol`)) +
  geom_point()+
  coord_flip()+
  xlab("Brain Structure") +
  ylab("Expression level (log2)") +
  ggtitle("RNA Expression of CYP450s in Different Brain Structures")

```
**Figure 3: Expression of CYP450s in different brain structures.** 

<br>
<br>


The average expression for each gene was calculated for each donor and brains structure, and compared to the average overall CYP11A1 expression. The genes with equal or higher expression than CYP11A1 were filtered out. The filtered data set was combined with the UniProt data to view more information about these genes. 
```{r, echo=FALSE}
averages <- unique(microarray[, .(
  gene_name = `gene-symbol`, 
  donor_age = donor_age,
  avg = mean(expression, na.rm = TRUE)
), by = c("gene-symbol","donor_id","top_level_structure_name")])
averages <- averages[order(-avg),]

cyp11a1_avg <- mean(averages[grep("CYP11A1", `gene-symbol`)]$avg, na.rm = TRUE)
averages_higher <- averages[avg >= cyp11a1_avg]

combined <- merge(averages_higher, cyp450, by.x = "gene_name", by.y = "Gene_name", all.x=TRUE, all.y=FALSE)

combined[, ER := grepl("endoplasmic\\sreticulum",Subcellular_location)]
combined[, Mitochondria := grepl("mitochondria|mitochondrion", Subcellular_location)]
combined[, Cytoplasm := grepl("cytoplasm|cytoplasmic", Subcellular_location)]
combined[, Organelle := ifelse(ER&Mitochondria, "Both ER and Mitochondria", 
                                  ifelse(Mitochondria, "Mitochondria Only", "ER Only"))]

combined$donor_age <- as.integer(stringr::str_remove_all(combined$donor_age, "[[:space:]]years"))
```

```{r, echo=FALSE}
combined %>%
  plot_ly(x=~avg, y=~top_level_structure_name, 
           type= 'scatter', mode = 'markers', color=~gene_name,
           hoverinfo='text',
           text=~paste(paste(gene_name, ":", sep=""), avg, paste("Subcellular location: ", Organelle), sep = "<br>")) %>%
 layout(title = "Expression of CYP450s with Higher or Equal Expression than CYP11A1 in the Brain",
                  xaxis = list(title = "Average expression (log2)"), yaxis = list(title = "Brain Structure"),
         hovermode = "compare")
```

**Figure 4: RNA Expression of Highly Expressed CYP450s at different ages.** There are `r length(unique(averages_higher$gene_name))` CYP450 enzymes with equal or higher expression than CYP11A1 in the brain. As seen from the graph, CYP51A1 appears to have the highest expression out of all the CYP450s, followed by CYP4X1 and CYP2J2. 

<br>
<br>


Now seeing which of these highly expressed CYP450s are involved in cholesterol or steroid function.  

```{r, echo=FALSE}
combined_chol <- combined[grep("cholesterol|steroid|hormone", Function),]

combined_chol2 <- unique(combined_chol[, .(
  gene_name = gene_name,
  Mass = Mass,
  Function = Function,
  Organelle = Organelle,
  Expression = mean(avg, na.rm = TRUE)
), by= "gene_name"])
combined_chol2 <- combined_chol2[order(-Expression),]

interest2 <- combined_chol2[, .(gene_name,Expression, Mass, Organelle)]

knitr::kable(interest2)
```

**Table 3: CYP450s with cholesterol related functions with RNA expression equal to or higher than CYP11A1 in the brain.** 
We see that there are `r length(unique(combined_chol$gene_name))` CYP450s with cholesterol related function when the two datasets are combined together, which are listed above. 

<br>
<br>

Finally, it would be interesting to see if the expression of these CYP450s change with age. Genes that do not have enough data points to show age-related trends were removed from the plots.  


```{r, echo=FALSE, warning =FALSE}
p<-ggplot(combined_chol[!grep("CYP2C9|CYP7A1", gene_name),], mapping=aes(x=donor_age, y = avg))+
  geom_point() +
  geom_smooth()+
  facet_wrap(vars(gene_name)) +
  xlab("Donor Age (years)") +
  ylab("Expression level (log2)") +
  ggtitle("Expression Change of Cholesterol/Steroid-related Brain CYP450s with Age")
ggplotly(p)

```
**Figure 5: Expression change of brain CYP450s involved in cholesterol/steroid function with age.** There appears to be no changes in the expressions of these CYP450s with age. 

<br>
<br>


## Conclusion
In this analysis, CYP450 enzymes that are expressed in the human brain with functions related to cholesterol and steroid were extracted from a data set containing all human CYP450s. There are `r nrow(chol_function)` of such proteins, which are `r chol_function$Gene_name`. Analysis of microarray data revealed `r length(unique(combined_chol$gene_name))` of such genes: `r interest2$gene_name`.

The expression for these genes in the brain can then be validated in future experiments using qRT-PCR and protein analysis on human brain tissue RNA. 

Full report can be viewed [here](https://github.com/yqchristina/PM566_Final_Project/blob/master/Report.pdf).

